---
title: "Cincinnati Reds"
subtitle: "Assessment 2.5"
author: "Sam Rizzuto"
date: "7 February 2022"
output:
  pdf_document:
    highlight: haddock
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{amsmath}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---

```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, # Required
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE # Turn off warnings
                      )

```

\clearpage



# Load Necessary Libraries
```{r}
library(dplyr)
library(mgcv)
library(parallel)
library(visreg)
library(ggplot2)
library(randomForest)
library(e1071)
library(caret)
library(flexclust)
library(factoextra)
library(knitr)

#set working directory
setwd("~/Desktop/22-ds")

#load in datasets
trainDF <- read.csv("train.csv")
testDF <- read.csv("test.csv")

```



# Filtering Data and Removing Outliers Through IQR
```{r}
Q1_Angle <- quantile(trainDF$ANGLE, .25)
Q3_Angle <- quantile(trainDF$ANGLE, .75)
IQR_Angle <- IQR(trainDF$ANGLE)

Q1_EXIT_SPEED <- quantile(trainDF$EXIT_SPEED, .25)
Q3_EXIT_SPEED <- quantile(trainDF$EXIT_SPEED, .75)
IQR_EXIT_SPEED <- IQR(trainDF$EXIT_SPEED)

Q1_DIRECTION <- quantile(trainDF$DIRECTION, .25)
Q3_DIRECTION <- quantile(trainDF$DIRECTION, .75)
IQR_DIRECTION <- IQR(trainDF$DIRECTION)

Q1_ReleaseSpeed <- quantile(trainDF$RELEASE_SPEED, .25)
Q3_ReleaseSpeed <- quantile(trainDF$RELEASE_SPEED, .75)
IQR_ReleaseSpeed <- IQR(trainDF$RELEASE_SPEED)

Q1_PlateX <- quantile(trainDF$PLATE_X, .25)
Q3_PlateX <- quantile(trainDF$PLATE_X, .75)
IQR_PlateX <- IQR(trainDF$PLATE_X)

Q1_PlateZ <- quantile(trainDF$PLATE_Z, .25)
Q3_PlateZ <- quantile(trainDF$PLATE_Z, .75)
IQR_PlateZ <- IQR(trainDF$PLATE_Z)

Q1_InducedVertBreak <- quantile(trainDF$INDUCED_VERTICAL_BREAK, .25)
Q3_InducedVertBreak <- quantile(trainDF$INDUCED_VERTICAL_BREAK, .75)
IQR_InducedVertBreak <- IQR(trainDF$INDUCED_VERTICAL_BREAK)

Q1_HorizontalBreak <- quantile(trainDF$HORIZONTAL_BREAK, .25)
Q3_HorizontalBreak <- quantile(trainDF$HORIZONTAL_BREAK, .75)
IQR_HorizontalBreak <- IQR(trainDF$HORIZONTAL_BREAK)

Q1_VertApproachAngle <- quantile(trainDF$VERTICAL_APPROACH_ANGLE, .25)
Q3_VertApproachAngle <- quantile(trainDF$VERTICAL_APPROACH_ANGLE, .75)
IQR_VertApproachAngle <- IQR(trainDF$VERTICAL_APPROACH_ANGLE)

Q1_HorizApproachAngle <- quantile(trainDF$HORIZONTAL_APPROACH_ANGLE, .25)
Q3_HorizApproachAngle <- quantile(trainDF$HORIZONTAL_APPROACH_ANGLE, .75)
IQR_HorizApproachAngle <- IQR(trainDF$HORIZONTAL_APPROACH_ANGLE)


trainDF <- subset(trainDF, trainDF$ANGLE > (Q1_Angle - 1.5*IQR_Angle) & 
                    trainDF$ANGLE < (Q3_Angle + 1.5*IQR_Angle))
trainDF <- subset(trainDF, trainDF$EXIT_SPEED > (Q1_EXIT_SPEED - 1.5*IQR_EXIT_SPEED) & 
                    trainDF$EXIT_SPEED < (Q3_EXIT_SPEED + 1.5*IQR_EXIT_SPEED))
trainDF <- subset(trainDF, trainDF$DIRECTION > (Q1_DIRECTION - 1.5*IQR_DIRECTION) & 
                    trainDF$DIRECTION < (Q3_DIRECTION + 1.5*IQR_DIRECTION))

trainDF <- subset(trainDF, trainDF$RELEASE_SPEED > (Q1_ReleaseSpeed - 1.5*IQR_ReleaseSpeed) & 
                    trainDF$RELEASE_SPEED < (Q3_ReleaseSpeed + 1.5*IQR_ReleaseSpeed))
trainDF <- subset(trainDF, trainDF$PLATE_X > (Q1_PlateX - 1.5*IQR_PlateX) & 
                    trainDF$PLATE_X < (Q3_PlateX + 1.5*IQR_PlateX))
trainDF <- subset(trainDF, trainDF$PLATE_Z > (Q1_PlateZ - 1.5*IQR_PlateZ) & 
                    trainDF$PLATE_Z < (Q3_PlateZ + 1.5*IQR_PlateZ))
trainDF <- subset(trainDF, trainDF$INDUCED_VERTICAL_BREAK > (Q1_InducedVertBreak - 1.5*IQR_InducedVertBreak) &
                    trainDF$INDUCED_VERTICAL_BREAK < (Q3_InducedVertBreak + 1.5*IQR_InducedVertBreak))
trainDF <- subset(trainDF, trainDF$HORIZONTAL_BREAK > (Q1_HorizontalBreak - 1.5*IQR_HorizontalBreak) &
                    trainDF$HORIZONTAL_BREAK < (Q3_HorizontalBreak + 1.5*IQR_HorizontalBreak))
trainDF <- subset(trainDF, trainDF$VERTICAL_APPROACH_ANGLE > (Q1_VertApproachAngle - 1.5*IQR_VertApproachAngle) &
                    trainDF$VERTICAL_APPROACH_ANGLE < (Q3_VertApproachAngle + 1.5*IQR_VertApproachAngle))
trainDF <- subset(trainDF, trainDF$HORIZONTAL_APPROACH_ANGLE > 
                    (Q1_HorizApproachAngle - 1.5*IQR_HorizApproachAngle) & 
                    trainDF$HORIZONTAL_APPROACH_ANGLE < (Q3_HorizApproachAngle + 1.5*IQR_HorizApproachAngle))

#disregard strikeouts, hbp, walks
trainDF <- trainDF %>% filter(PITCH_RESULT_KEY == "InPlay")
#goes from original 26417 rows to 24510 after cleaning

summary(trainDF)
######################

#create ops and handedness variables
trainDF <- trainDF %>% mutate(OPS = OBP + SLG)
testDF <- testDF %>% mutate(OPS = OBP + SLG)
trainDF <- trainDF %>% mutate(Handedness = if_else(THROWS_LEFT == BATS_LEFT, 0, 1))
testDF <- testDF %>% mutate(Handedness = if_else(THROWS_LEFT == BATS_LEFT, 0, 1))
#0 if same hands facing eachother, 1 if opposite
```

# Random Forest Model
```{r}
#running random forest on 3 prediction variables
rfTrainExitVelo <- randomForest(EXIT_SPEED ~ RELEASE_SPEED + PLATE_X + PLATE_Z + 
                                  INDUCED_VERTICAL_BREAK + HORIZONTAL_BREAK +
                                  VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + 
                                  OPS + Handedness, data = trainDF)
rfTrainAngle <- randomForest(ANGLE ~ RELEASE_SPEED + PLATE_X + PLATE_Z + 
                               INDUCED_VERTICAL_BREAK + HORIZONTAL_BREAK +
                               VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + 
                               OPS + Handedness, data = trainDF)
rfTrainDirection <- randomForest(DIRECTION ~ RELEASE_SPEED + PLATE_X + PLATE_Z + 
                                   INDUCED_VERTICAL_BREAK + HORIZONTAL_BREAK +
                                   VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + 
                                   OPS + Handedness, data = trainDF)
#viewing importance plots of each model type to determine most significant vars in model
varImpPlot(rfTrainExitVelo) #drop handedness
varImpPlot(rfTrainAngle) #drop ops, handedness
varImpPlot(rfTrainDirection) #drop handedness

#choose first grouping vars
rfUpdatedExitVelo <- randomForest(EXIT_SPEED ~ PLATE_X + PLATE_Z + HORIZONTAL_APPROACH_ANGLE +
                                    HORIZONTAL_BREAK + RELEASE_SPEED +
                                    VERTICAL_APPROACH_ANGLE + INDUCED_VERTICAL_BREAK + OPS, data = trainDF)
importance(rfUpdatedExitVelo) #importance vars of rf exit velo model
plot(rfUpdatedExitVelo) #error of rf exit velo model

rfUpdatedAngle <- randomForest(ANGLE ~ PLATE_Z + INDUCED_VERTICAL_BREAK + RELEASE_SPEED +
                                 VERTICAL_APPROACH_ANGLE +
                                PLATE_X + HORIZONTAL_BREAK + HORIZONTAL_APPROACH_ANGLE , data = trainDF)

rfUpdatedDirection <- randomForest(DIRECTION ~  PLATE_X + HORIZONTAL_APPROACH_ANGLE + RELEASE_SPEED + 
                                     HORIZONTAL_BREAK + VERTICAL_APPROACH_ANGLE +
                                     INDUCED_VERTICAL_BREAK + PLATE_Z + OPS, data = trainDF)



#plot randomForest results back in trainDF
trainDF <- trainDF %>% 
  mutate(exitVelo_RF = round(predict(rfUpdatedExitVelo, newdata = .),3))
trainDF <- trainDF %>% 
  mutate(angle_RF = round(predict(rfUpdatedAngle, newdata = .),3))
trainDF <- trainDF %>% 
  mutate(direction_RF = round(predict(rfUpdatedDirection, newdata = .),3))

```

# SVM Model
```{r}
#running svm on exit velo
svmExitVelo <- svm(EXIT_SPEED ~ RELEASE_SPEED + PLATE_X + PLATE_Z + 
                     INDUCED_VERTICAL_BREAK + HORIZONTAL_BREAK +
                     VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + 
                     OPS + Handedness, data = trainDF, cost = 100, gamma = 1)
#removing predictor variable of exit velo
svmExitVelo_Pred <- round(predict(svmExitVelo, trainDF[,-19]), 3)
#add svm exit velo into training df
trainDF <- trainDF %>% mutate(exitVelo_SVM = svmExitVelo_Pred)



#running svm on angle
svmAngle <- svm(ANGLE ~ RELEASE_SPEED + PLATE_X + PLATE_Z + 
                  INDUCED_VERTICAL_BREAK + HORIZONTAL_BREAK +
                  VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + 
                  OPS + Handedness, data = trainDF, cost = 100, gamma = 1)
#removing predictor variable of angle
svmAngle_Pred <- round(predict(svmAngle, trainDF[,-20]), 3)
#adding prob of angle into training df
trainDF <- trainDF %>% mutate(angle_SVM = svmAngle_Pred)



#running svm model on direction
svmDirection <- svm(DIRECTION ~ RELEASE_SPEED + PLATE_X + PLATE_Z + 
                      INDUCED_VERTICAL_BREAK + HORIZONTAL_BREAK +
                      VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + 
                      OPS + Handedness, data = trainDF, cost = 100, gamma = 1)
#removing predictor variable of direction
svmDirection_Pred <- round(predict(svmDirection, trainDF[,-21]), 3)
#adding direction into training df
trainDF <- trainDF %>% mutate(direction_SVM = svmDirection_Pred)

```

# Running GAM Model
```{r}

####generalized additive model to include most important variables in training dataset
options(mc.cores = parallel::detectCores())#run model in parallel

gam_EV <- bam(EXIT_SPEED ~ RELEASE_SPEED + PLATE_X + PLATE_Z + INDUCED_VERTICAL_BREAK +
              HORIZONTAL_BREAK + VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
            data = trainDF, family = gaussian, method = "GCV.Cp")
summary(gam_EV) #drop platex, platez, vert approach angle
gam_EV_Upd <- bam(EXIT_SPEED ~ RELEASE_SPEED + INDUCED_VERTICAL_BREAK +
                HORIZONTAL_BREAK + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
              data = trainDF, family = gaussian, method = "GCV.Cp")

gam_Ang <- bam(ANGLE ~ RELEASE_SPEED + PLATE_X + PLATE_Z + INDUCED_VERTICAL_BREAK + 
              HORIZONTAL_BREAK + VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
            data = trainDF, family = gaussian, method = "GCV.Cp")
summary(gam_Ang) #no drop
gam_Ang_Upd <- gam_Ang

gam_Dir <- bam(DIRECTION ~ RELEASE_SPEED + PLATE_X + PLATE_Z + INDUCED_VERTICAL_BREAK +
              HORIZONTAL_BREAK + VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
            data = trainDF, family = gaussian, method = "GCV.Cp")
summary(gam_Dir) #drop platez
gam_Dir_Upd <- bam(DIRECTION ~ RELEASE_SPEED + PLATE_X + INDUCED_VERTICAL_BREAK +
                 HORIZONTAL_BREAK + VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
               data = trainDF, family = gaussian, method = "GCV.Cp")

#coefficients of each model to use in prediciting values
gamEV <- gam_EV_Upd$coefficients
gamAng <- gam_Ang_Upd$coefficients
gamDir <- gam_Dir_Upd$coefficients

trainDF <- trainDF %>% 
  mutate(exitVelo_GAM = round(gamEV[1] + gamEV[2]*RELEASE_SPEED + gamEV[3]*INDUCED_VERTICAL_BREAK + 
                                gamEV[4]*HORIZONTAL_BREAK + gamEV[5]*HORIZONTAL_APPROACH_ANGLE + 
                                gamEV[6]*OPS + gamEV[7]*Handedness, 3))

trainDF <- trainDF %>% 
  mutate(angle_GAM = round(gamAng[1] + gamAng[2]*RELEASE_SPEED + gamAng[3]*PLATE_X + 
                             gamAng[4]*PLATE_Z + gamAng[5]*INDUCED_VERTICAL_BREAK + 
                             gamAng[6]*HORIZONTAL_BREAK + gamAng[7]*VERTICAL_APPROACH_ANGLE +
                             gamAng[8]*HORIZONTAL_APPROACH_ANGLE + 
                             gamAng[9]*OPS + gamAng[10]*Handedness, 3))

trainDF <- trainDF %>% 
  mutate(direction_GAM = round(gamDir[1] + gamDir[2]*RELEASE_SPEED + gamDir[3]*PLATE_X + 
                                 gamDir[4]*INDUCED_VERTICAL_BREAK +gamDir[5]*HORIZONTAL_BREAK +
                                 gamDir[6]*VERTICAL_APPROACH_ANGLE + 
                                 gamDir[7]*HORIZONTAL_APPROACH_ANGLE + gamDir[8]*OPS + 
                                 gamDir[9]*Handedness, 3))

```

# Smaller GAM Model to Create Visuals
```{r}
###smaller gam
#exit speed just on x and y location
gam_Small <- bam(EXIT_SPEED ~ s(PLATE_X, PLATE_Z), 
            data = trainDF, family = gaussian, method = "GCV.Cp")

xs <- matrix(data=seq(from=-2, to=2, length=50), nrow=50, ncol=50)    
ys <- t(matrix(data=seq(from=0,to=5, length=50), nrow=50, ncol=50))

gamSmallFit <- data.frame(PLATE_X = as.vector(xs), PLATE_Z = as.vector(ys))
exitVeloPred <- predict(gam_Small, gamSmallFit, types = "response")
exitVeloPred <- matrix(exitVeloPred, nrow = 50, ncol = 50)
density(exitVeloPred)
summary(trainDF$EXIT_SPEED)
summary(trainDF$PLATE_X) #-1.5min to 1.5max, so round to -2 and 2 for x
summary(trainDF$PLATE_Z) #0.8 to 4.9, so round to 0 and 5
#range of 60-110 to get full exit velo based on iqr
#width of HP is 23in 
#height of strikezone is 25.79in based on baseball prospectus,
#with the bottom starting at 18.29in above the ground, which is 1.52ft
#to get the top of the zone, add 1.52ft plus the height (25.79in) to get 44.08in, or 3.67ft
#To create width of strikezone (17in), divide by 2 = 8.5in on each side from the middle
#and convert to ft = 0.71ft from the center
#so now when drawing strikezone, it has width -0.71 to 0.71 and height 1.52 to 3.67
filled.contour(x=seq(from=-2, to=2, length=50), y=seq(from=0, to=5, length=50), z = exitVeloPred, 
               zlim=c(60,110), 
               color.palette = colorRampPalette(c("lightblue","yellow","orange", "red", "darkred")),
               plot.axes = { rect(-0.71, 1.52, 0.71, 3.67, border="black", lty="dashed", lwd=3)
                 axis(1, at=c(-2,-1,0,1,2), pos=0, labels=c(-2,-1,0,1,2), las=0, col="black")
                 axis(2, at=c(0,1,2,3,4,5), pos=-2, labels=c(0,1,2,3,4,5), las=0, col="black")
                 }, 
               main = "Heat Map for Exit Velo Based on \n X/Y Location of Strikezone", 
               ylab = "Vertical (ft)", 
               xlab = "Horizontal Location of Pitch at Plate (ft)")

#launch angle
gam_Small2 <- bam(ANGLE ~ s(PLATE_X, PLATE_Z), 
                 data = trainDF, family = gaussian, method = "GCV.Cp")
gamSmallFit2 <- data.frame(PLATE_X = as.vector(xs), PLATE_Z = as.vector(ys))
anglePred <- predict(gam_Small2, gamSmallFit2, types = "response")
anglePred <- matrix(anglePred, nrow = 50, ncol = 50)
summary(trainDF$ANGLE)
filled.contour(x=seq(from=-2, to=2, length=50), y=seq(from=0, to=5, length=50), z = anglePred, 
               zlim=c(-15,35), 
               color.palette = colorRampPalette(c("lightblue","yellow","orange", "red", "darkred")),
               plot.axes = { rect(-0.71, 1.52, 0.71, 3.67, border="black", lty="dashed", lwd=3)
                 axis(1, at=c(-2,-1,0,1,2), pos=0, labels=c(-2,-1,0,1,2), las=0, col="black")
                 axis(2, at=c(0,1,2,3,4,5), pos=-2, labels=c(0,1,2,3,4,5), las=0, col="black")
               }, 
               main = "Heat Map for Launch Angle Based on \n X/Y Location of Strikezone", 
               ylab = "Vertical (ft)", 
               xlab = "Horizontal Location of Pitch at Plate (ft)")


#direction
gam_Small3 <- bam(DIRECTION ~ s(PLATE_X, PLATE_Z), 
                  data = trainDF, family = gaussian, method = "GCV.Cp")
gamSmallFit3 <- data.frame(PLATE_X = as.vector(xs), PLATE_Z = as.vector(ys))
directionPred <- predict(gam_Small3, gamSmallFit3, types = "response")
directionPred <- matrix(directionPred, nrow = 50, ncol = 50)
summary(trainDF$DIRECTION)
#not taking into account handedness (l/r are combined here)
filled.contour(x=seq(from=-2, to=2, length=50), y=seq(from=0, to=5, length=50), z = directionPred, 
               zlim=c(-25,25), 
               color.palette = colorRampPalette(c("lightblue","yellow","orange", "red", "darkred")),
               plot.axes = { rect(-0.71, 1.52, 0.71, 3.67, border="black", lty="dashed", lwd=3)
                 axis(1, at=c(-2,-1,0,1,2), pos=0, labels=c(-2,-1,0,1,2), las=0, col="black")
                 axis(2, at=c(0,1,2,3,4,5), pos=-2, labels=c(0,1,2,3,4,5), las=0, col="black")
               }, 
               main = "Heat Map for Direction Based on \n X/Y Location of Strikezone", 
               ylab = "Vertical (ft)", 
               xlab = "Horizontal Location of Pitch at Plate (ft)")






#building predicted values with upper/lower bounds
exitVeloPredConf <- predict(gam_Small, gamSmallFit, types = "response", se = TRUE)
predVals <- data.frame(gamSmallFit, exitVeloPredConf) %>% 
  mutate(lower = exitVeloPredConf$fit - 1.96*exitVeloPredConf$se.fit,
         upper = exitVeloPredConf$fit + 1.96*exitVeloPredConf$se.fit)

ggplot(aes(x=PLATE_X,y=exitVeloPredConf$fit), data=predVals) +
  geom_ribbon(aes(ymin = lower, ymax=upper), fill='gray90') +
  geom_line(color='#00aaff') + ylab("Exit Velocity Prediction") + 
  ggtitle("Prediction of Exit Velocity Based on Horizontal Location of Pitch")
ggplot(aes(x=PLATE_Z,y=exitVeloPredConf$fit), data=predVals) +
  geom_ribbon(aes(ymin = lower, ymax=upper), fill='gray90') +
  geom_line(color='#00aaff') + ylab("Exit Velocity Prediction") + 
  ggtitle("Prediction of Exit Velocity Based on Vertical Location of Pitch")

#both plots show where the highest exit velo is based on x and y location of pitch

vis.gam(gam_Small, type='response', plot.type='contour', main = "Exit Velo")
visreg2d(gam_Small, xvar='PLATE_X', yvar='PLATE_Z', scale='response')

visreg2d(gam_EV_Upd, xvar='RELEASE_SPEED', yvar='OPS', scale='response')

anova(gam_EV_Upd, gam_Small, test="Chisq")
anova(gam_EV_Upd)
anova(gam_Ang_Upd)
anova(gam_Dir_Upd)

```

# GLM model
```{r}
############exit velo
exitVeloCalc <- glm(EXIT_SPEED ~ RELEASE_SPEED + PLATE_X + PLATE_Z + INDUCED_VERTICAL_BREAK + 
                      HORIZONTAL_BREAK +
                      VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
                  data = trainDF, 
                  family = gaussian)
summary(exitVeloCalc) 
exitVeloCalc$coefficients #coefficients used to determine exit velo
anova(exitVeloCalc, test = "Chisq")

############angle
angleCalc <- glm(ANGLE ~ RELEASE_SPEED + PLATE_X + PLATE_Z + INDUCED_VERTICAL_BREAK + HORIZONTAL_BREAK +
                      VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
                    data = trainDF, 
                    family = gaussian)
anova(angleCalc, test = "Chisq")


############direction
directionCalc <- glm(DIRECTION ~ RELEASE_SPEED + PLATE_X + PLATE_Z + INDUCED_VERTICAL_BREAK + 
                       HORIZONTAL_BREAK +
                      VERTICAL_APPROACH_ANGLE + HORIZONTAL_APPROACH_ANGLE + OPS + Handedness, 
                    data = trainDF, 
                    family = gaussian)
anova(directionCalc, test = "Chisq")

evc <- exitVeloCalc$coefficients
ac <- angleCalc$coefficients
dc <- directionCalc$coefficients

#adding new glm predicted columns to training df
trainDF <- trainDF %>%
  mutate(exitVelo_GLM = round(evc[1] + evc[2]*RELEASE_SPEED + evc[3]*PLATE_X + evc[4]*PLATE_Z +
                                evc[5]*INDUCED_VERTICAL_BREAK + evc[6]*HORIZONTAL_BREAK +
                                evc[7]*VERTICAL_APPROACH_ANGLE + 
                                evc[8]*HORIZONTAL_APPROACH_ANGLE + evc[9]*OPS + evc[10]*Handedness, 3))


trainDF <- trainDF %>%
  mutate(angle_GLM = round(ac[1] + ac[2]*RELEASE_SPEED + ac[3]*PLATE_X + ac[4]*PLATE_Z +
                             ac[5]*INDUCED_VERTICAL_BREAK + ac[6]*HORIZONTAL_BREAK +
                             ac[7]*VERTICAL_APPROACH_ANGLE + 
                             ac[8]*HORIZONTAL_APPROACH_ANGLE + ac[9]*OPS + ac[10]*Handedness, 3))

trainDF <- trainDF %>%
  mutate(direction_GLM = round(dc[1] + dc[2]*RELEASE_SPEED + dc[3]*PLATE_X + dc[4]*PLATE_Z +
                                 dc[5]*INDUCED_VERTICAL_BREAK + dc[6]*HORIZONTAL_BREAK +
                                 dc[7]*VERTICAL_APPROACH_ANGLE + 
                                 dc[8]*HORIZONTAL_APPROACH_ANGLE + dc[9]*OPS + dc[10]*Handedness, 3))

```

# KMeans
```{r}
subsetTrain <- trainDF[,c(19, 12:18, 29:30)]
km1 <- kmeans(subsetTrain[,2:10], 3, iter.max = 100)

clusters <- factor(km1$cluster)
ggplot(trainDF, aes(x = EXIT_SPEED, y = RELEASE_SPEED, color = clusters)) + geom_point() +
  ggtitle("Release Speed vs Exit Velocity Grouped By Clusters")


km2 <- kmeans(scale(trainDF[, c(12:18, 29:30)]), 3, nstart = 25, iter.max = 100)
fviz_cluster(km2, data = trainDF[, c(12:18, 29:30)], 
             main = "KMeans Plot By Clusters Based on Var Similarity")


# Dimension reduction PCA
pca <- prcomp(trainDF[, c(12:18, 29:30)],  scale = TRUE)
cords <- as.data.frame(get_pca_ind(pca)$coord)
cords$cluster1 <- factor(km2$cluster)
cords$EXIT_SPEED <- trainDF$EXIT_SPEED
head(cords) #see first few rows dimensions and clusters

```

# KNN
```{r}
#######knn on exit velo
trainKNN_x_EV <- trainDF[,c(12:18,29:30)]
trainKNN_y_EV <- trainDF[,19]
knnModel_EV <- knnreg(trainKNN_x_EV, trainKNN_y_EV)
testKNN_x_EV <- testDF[,c(12:20)]

#testdf pred vals
knnpred_y_EV <- predict(knnModel_EV, data.frame(testKNN_x_EV))

knnpred_xTconf_EV <- predict(knnModel_EV, data.frame(trainKNN_x_EV), interval = "confidence", level = 0.9)
ggplot(trainDF, aes(EXIT_SPEED, knnpred_xTconf_EV)) + geom_smooth() +
  ggtitle("Error Band on KNN Model Predicting Exit Velocity") + ylab("Exit Velo Prediction")
  

mse <- mean((trainKNN_y_EV - knnpred_xTconf_EV)^2)
mae <- MAE(trainKNN_y_EV, knnpred_xTconf_EV)
rmse <- RMSE(trainKNN_y_EV, knnpred_xTconf_EV)
mse
mae
rmse



#####knn on angle
trainKNN_x_ang <- trainDF[,c(12:18,29:30)]
trainKNN_y_ang <- trainDF[,20]
knnModel_ang <- knnreg(trainKNN_x_ang, trainKNN_y_ang)
testKNN_x_ang <- testDF[,c(12:20)]

knnpred_y_ang <- predict(knnModel_ang, data.frame(testKNN_x_ang))

knnpred_xTconf_ang <- predict(knnModel_ang, data.frame(trainKNN_x_ang), interval = "confidence", level = 0.9)
ggplot(trainDF, aes(ANGLE, knnpred_xTconf_ang)) + geom_smooth() +
  ggtitle("Error Band on KNN Model Predicting Launch Angle") + ylab("Angle Prediction")

#####knn on direction
trainKNN_x_dir <- trainDF[,c(12:18,29:30)]
trainKNN_y_dir <- trainDF[,21]
knnModel_dir <- knnreg(trainKNN_x_dir, trainKNN_y_dir)
testKNN_x_dir <- testDF[,c(12:20)]

knnpred_y_dir <- predict(knnModel_dir, data.frame(testKNN_x_dir))

knnpred_xTconf_dir <- predict(knnModel_dir, data.frame(trainKNN_x_dir), interval = "confidence", level = 0.9)
ggplot(trainDF, aes(DIRECTION, knnpred_xTconf_dir)) + geom_smooth() +
  ggtitle("Error Band on KNN Model Predicting Direction") + ylab("Direction Prediction")

#add pred vals to training df
trainDF <- trainDF %>% mutate(exitVelo_KNN = round(knnpred_xTconf_EV,3))
trainDF <- trainDF %>% mutate(angle_KNN = round(knnpred_xTconf_ang,3))
trainDF <- trainDF %>% mutate(direction_KNN = round(knnpred_xTconf_dir,3))
```

# Finding Absolute Errors on Training DF Models
```{r}
#######################
mean(abs(trainDF$EXIT_SPEED - trainDF$exitVelo_RF))
mean(abs(trainDF$EXIT_SPEED - trainDF$exitVelo_SVM))
mean(abs(trainDF$EXIT_SPEED - trainDF$exitVelo_GAM))
mean(abs(trainDF$EXIT_SPEED - trainDF$exitVelo_GLM))
mean(abs(trainDF$EXIT_SPEED - trainDF$exitVelo_KNN))

mean(abs(trainDF$ANGLE - trainDF$angle_RF))
mean(abs(trainDF$ANGLE - trainDF$angle_SVM))
mean(abs(trainDF$ANGLE - trainDF$angle_GAM))
mean(abs(trainDF$ANGLE - trainDF$angle_GLM))
mean(abs(trainDF$ANGLE - trainDF$angle_KNN))

mean(abs(trainDF$DIRECTION - trainDF$direction_RF))
mean(abs(trainDF$DIRECTION - trainDF$direction_SVM))
mean(abs(trainDF$DIRECTION - trainDF$direction_GAM))
mean(abs(trainDF$DIRECTION - trainDF$direction_GLM))
mean(abs(trainDF$DIRECTION - trainDF$direction_KNN))
######################
```
It can be seen that SVM has the smallest mean error rate

# Testing Data
```{r}
##################Testing Data
####gam
#adding gam predicted values to test df
testDF <- testDF %>% 
  mutate(exitVelo_GAM = round(gamEV[1] + gamEV[2]*RELEASE_SPEED + gamEV[3]*INDUCED_VERTICAL_BREAK + 
                                gamEV[4]*HORIZONTAL_BREAK + gamEV[5]*HORIZONTAL_APPROACH_ANGLE + 
                                gamEV[6]*OPS + gamEV[7]*Handedness, 3))

testDF <- testDF %>% 
  mutate(angle_GAM = round(gamAng[1] + gamAng[2]*RELEASE_SPEED + gamAng[3]*PLATE_X + 
                             gamAng[4]*PLATE_Z + gamAng[5]*INDUCED_VERTICAL_BREAK + 
                             gamAng[6]*HORIZONTAL_BREAK + gamAng[7]*VERTICAL_APPROACH_ANGLE + 
                             gamAng[8]*HORIZONTAL_APPROACH_ANGLE + 
                             gamAng[9]*OPS + gamAng[10]*Handedness, 3))

testDF <- testDF %>% 
  mutate(direction_GAM = round(gamDir[1] + gamDir[2]*RELEASE_SPEED + gamDir[3]*PLATE_X + 
                                 gamDir[4]*INDUCED_VERTICAL_BREAK + gamDir[5]*HORIZONTAL_BREAK + 
                                 gamDir[6]*VERTICAL_APPROACH_ANGLE + 
                                 gamDir[7]*HORIZONTAL_APPROACH_ANGLE + 
                                 gamDir[8]*OPS + gamDir[9]*Handedness, 3))

####glm
#adding glm predicted values to test df
testDF <- testDF %>% 
  mutate(exitVelo_GLM = round(evc[1] + evc[2]*RELEASE_SPEED + evc[3]*PLATE_X + evc[4]*PLATE_Z +
                                evc[5]*INDUCED_VERTICAL_BREAK + evc[6]*HORIZONTAL_BREAK +
                                evc[7]*VERTICAL_APPROACH_ANGLE + 
                                evc[8]*HORIZONTAL_APPROACH_ANGLE + evc[9]*OPS + 
                                evc[10]*Handedness, 3))


testDF <- testDF %>% 
  mutate(angle_GLM = round(ac[1] + ac[2]*RELEASE_SPEED + ac[3]*PLATE_X + ac[4]*PLATE_Z +
                             ac[5]*INDUCED_VERTICAL_BREAK + ac[6]*HORIZONTAL_BREAK +
                             ac[7]*VERTICAL_APPROACH_ANGLE + 
                             ac[8]*HORIZONTAL_APPROACH_ANGLE + ac[9]*OPS + 
                             ac[10]*Handedness, 3))

testDF <- testDF %>% 
  mutate(direction_GLM = round(dc[1] + dc[2]*RELEASE_SPEED + dc[3]*PLATE_X + dc[4]*PLATE_Z +
                                 dc[5]*INDUCED_VERTICAL_BREAK + dc[6]*HORIZONTAL_BREAK +
                                 dc[7]*VERTICAL_APPROACH_ANGLE + 
                                 dc[8]*HORIZONTAL_APPROACH_ANGLE + dc[9]*OPS + 
                                 dc[10]*Handedness, 3))

###knn
#adding knn predicted values to test df
testDF <- testDF %>% mutate(exitVelo_KNN = round(knnpred_y_EV,3))
testDF <- testDF %>% mutate(angle_KNN = round(knnpred_y_ang,3))
testDF <- testDF %>% mutate(direction_KNN = round(knnpred_y_dir,3))

#View first few rows of my altered df's to see my predicted values for each model
head(trainDF)
head(testDF)

#writing to working directory to display final result tables
# write.csv(trainDF, "myTrainDF.csv")
# write.csv(testDF, "myTestDF.csv")

```


# Another approach I though of to run a simulation (but did not get to because of time limit) is this:
Find the distribution of each of the 3 vars based on pitch type (say there are 4 pitch types) from k means and using the batting/pitch info, repeatedly generate outcome result (from model in assessment 2) for each var based on each pitch type and then randomly grab a exit velo, LA, and direction and pull the assigned outcome (out, 1b, 2b, 3b, hr) then randomly runs 1000 times, take the avg, and assign that average to be the exit velo, LA, and direction, respectively, on test data
```{r, eval=FALSE}

smallSmall <- trainDF[1:5,12:44]

store2 <- list()
for (i in 1:nrow(smallSmall)) {
  storeTest <- sample(smallSmall[i,30:33], 1, prob = abs(c(smallSmall[i,30], 
                                                           smallSmall[i,31], 
                                                           smallSmall[i,32], 
                                                           smallSmall[i,33])))
  store2[[i]] <- storeTest
}
store2



trainDF <- trainDF %>% mutate(swingSpeed = EXIT_SPEED - (0.2 * trainDF$RELEASE_SPEED) / (1 + 0.2))
newDF <- trainDF %>% group_by(BATTER_UID) %>% summarise(avgSS = mean(swingSpeed),
                                                        n = n())
newDF %>% arrange(desc(avgSS))

hist(newDF$avgSS)

#0.2 is value of wooden 
#avg 70mph bat speed according to blast motion
#runif(nrow(trainDF), min = 40, max = 80) represents randomly assigning bat speed
exitVeloFormula <- (0.2 * trainDF$RELEASE_SPEED) + (1 + 0.2)*runif(nrow(trainDF), min = 40, max = 80)
summary(exitVeloFormula)
round(asin(trainDF$PLATE_X / exitVeloFormula*(21.922)),3)

```


